An example exploratory analysis script

### 2025-03-19 code/analysis-code/statistical-analysis.qdm.qmd
Data Analysis (Part 3 of the project)

install.packages("gridExtra")  # If not installed
library(gridExtra)
```

@@

@@

@@

# Define independent variables (exclude 'tbi' and categorical variables)
independent_vars <- colnames(asia_africa)[!colnames(asia_africa) %in% c("tbi", "country", "region", "year")]

# Create scatter plots for each independent variable against tbi
plots <- lapply(independent_vars, function(var) {
  ggplot(asia_africa, aes_string(x = var, y = "tbi")) +
    geom_point() + 
    ggtitle(paste("TBI vs", var)) + 
    xlab(var) + 
    ylab("TBI") + 
    theme_minimal()
})

# Create a list of scatter plots for each independent variable against tbi
plots <- lapply(independent_vars, function(var) {
  ggplot(asia_africa, aes_string(x = var, y = "tbi")) +
    geom_point() + 
    ggtitle(paste("tbi vs", var)) + 
    xlab(var) + 
    ylab("TBI") + 
    theme_minimal()
})

# Combine all plots into a single frame
combined_plot <- wrap_plots(plots, ncol = 3)  # Adjust ncol to the number of columns you want

combineplot_location <- here( "code","analysis-code", "figures", "scatterplot.png") # to set up location for the pictures created 
ggsave(filename = combineplot_location, plot=combined_plot, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
plot(combined_plot)

@@

@@

# Reshape the dataset to long format
asia_africa_long <- asia_africa %>%
  pivot_longer(cols = c(pro_cleanfuel, pro_basicwater, pro_safewater, 
                        mean_exposure_pm2.5, total_greenhouses, pro_basicsanitation, 
                        pro_safesanitation, gdp, hdi),
               names_to = "Predictor", values_to = "Value")

@@

Data transformation

@@

asia_africa <- asia_africa %>%
  mutate(
    log_tbi = log(tbi + 1),
    log_gdp = log(gdp + 1),
    log_total_greenhouses = log(total_greenhouses + 1),
    log_pm2.5 = log(mean_exposure_pm2.5 + 1),
    sqrt_basicwater = sqrt(pmax(pro_basicwater, 0)), 
    sqrt_safewater = sqrt(pmax(pro_safewater, 0)),
    poly_cleanfuel = as.numeric(poly(pro_cleanfuel, 2, raw = TRUE)[, 1]),
    poly_basicsanitation = as.numeric(poly(pro_basicsanitation, 2, raw = TRUE)[, 1]),
    poly_safesanitation = as.numeric(poly(pro_safesanitation, 2, raw = TRUE)[, 1])
  )

@@

```{r}
asia_africa_transformed <- asia_africa %>%
  dplyr::select(log_tbi, log_gdp, log_total_greenhouses, log_pm2.5, 
                poly_cleanfuel, poly_basicsanitation, poly_safesanitation,
                sqrt_basicwater, sqrt_safewater, hdi) %>%
  pivot_longer(cols = -log_tbi, names_to = "Predictor", values_to = "Value")

@@

>>>>

install.packages("gridExtra")  # If not installed
library(gridExtra)

install.packages("lmerTest")
library(lmerTest)
install.packages("stargazer")
library(stargazer)

@@

# In the study, I will focus on Water and Sanitation and remove Lesotho as it considered as outlier
asia_africa <- asia_africa%>%
  filter(country != "Lesotho") %>% #Drop Lesotho from the observation
  select("tbi", "year", "pro_basicwater", "pro_safewater", "pro_basicsanitation", "pro_safesanitation", "hdi", "country", "region")

@@

```{r}
# Adding Fixed-Effect to Make Conditional Models ---
ggplot(asia_africa, aes(x = year, y = tbi)) + 
  geom_line(aes(group=country), col="grey") + 
  stat_smooth(method="loess", col="black", lwd=1.5, se=FALSE)+
  scale_x_continuous(name = "years") + 
  scale_y_continuous(name = "Tuberculosis Incidence") +
  theme_bw() + 
  theme(axis.text=element_text(size=14, colour="black"), 
        axis.title=element_text(size=14,face="bold")) +
  theme(strip.text.x = element_text(size = 14))+ theme(legend.position="none") 
```

```{r}
# Convert dataset to long format for faceting
asia_africa_long <- asia_africa %>%
  pivot_longer(cols = c(pro_basicwater, pro_safewater,
                        pro_basicsanitation, pro_safesanitation, hdi),
               names_to = "variable", values_to = "value")

# Create scatter plot faceted by independent variable
scatter_faceted <- ggplot(asia_africa_long, aes(x = value, y = tbi, color = country)) + 
  geom_point(alpha = 0.6) +  # Scatter points, colored by country
  stat_smooth(method = "loess", col = "black", lwd = 1.2, se = FALSE) +  # LOESS (non-linear) in black
  facet_wrap(~ variable, scales = "free_x") +  # Facet by variable
  labs(x = "Independent Variable Value", y = "Tuberculosis Incidence (TBI)") +
  theme_bw() + 
  theme(axis.text = element_text(size = 12, colour = "black"), 
        axis.title = element_text(size = 14, face = "bold"),
        strip.text = element_text(size = 14, face = "bold"), 
        legend.position = "none")  # Hides the legend

nonrelationshiplot_location <- here( "code","analysis-code", "figures", "nonlinear.png") # to set up location for the pictures created 
ggsave(filename = nonrelationshiplot_location, plot=scatter_faceted, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
# Show plot
print(scatter_faceted)

Non-linearity test 
```{r}
# Install package if not already installed
install.packages("lmtest")
library(lmtest)

# Fit a linear model
lm_model <- lm(tbi ~ pro_cleanfuel + pro_basicwater + pro_safewater + 
               mean_exposure_pm2.5 + total_greenhouses + 
               pro_basicsanitation + pro_safesanitation + gdp + hdi, 
               data = asia_africa)

# Run the RESET test for linearity
resettest(lm_model, power = 2:3)

@@

# Fit a linear model (LM)
lm_model <- lm(tbi ~ pro_cleanfuel + pro_basicwater + pro_safewater + 
               mean_exposure_pm2.5 + total_greenhouses + 
               pro_basicsanitation + pro_safesanitation + gdp + hdi, 
               data = asia_africa)

# Fit a non-linear model (GAM)
gam_model <- gam(tbi ~ s(pro_cleanfuel) + s(pro_basicwater) + s(pro_safewater) + 
                  s(mean_exposure_pm2.5) + s(total_greenhouses) + 
                  s(pro_basicsanitation) + s(pro_safesanitation) + 
                  s(gdp) + s(hdi), 
                 data = asia_africa, method = "REML")

# Compare AIC values
AIC(lm_model, gam_model)

# Plot residuals vs. fitted values for the linear model
plot(lm_model$fitted.values, resid(lm_model), 
     xlab = "Fitted Values", ylab = "Residuals", 
     main = "Residuals vs. Fitted Plot")
abline(h = 0, col = "red")  # Horizontal line at 0

@@

```{r}
gamm_model <-gam(tbi ~ s(year) + s(pro_cleanfuel) + s(hdi) + ti(pro_cleanfuel, hdi) + s(country, bs="re"), 
    family = nb(), method = "REML", data = asia_africa)

```{r}
library(car)
vif(gamm_model)
```

```{r}
cor_matrix <- cor(asia_africa[, c("year", "pro_cleanfuel", "hdi", "gdp")], use="pairwise.complete.obs")
print(cor_matrix)

### Longitudinal analysis with Mixed-effect models
```{r}
# Pastikan tidak ada NA dalam variabel yang dianalisis
asia_africa_clean <- asia_africa %>%
  drop_na(tbi, pro_cleanfuel, mean_exposure_pm2.5, pro_basicwater, pro_safewater, pro_basicsanitation, pro_safesanitation, total_greenhouses, gdp, hdi, year, country)

# Model Mixed-Effects: TBI is predicted by predictors with random effect per each country 
model_lmm <- lmer(tbi ~ pro_cleanfuel+ mean_exposure_pm2.5 + pro_basicwater + pro_safewater + pro_basicsanitation + pro_safesanitation + total_greenhouses + gdp + hdi + year + (1 | country),  # Random intercept dressed variation in each country
                   data = asia_africa_clean, REML = FALSE)

# Ringkasan hasil model
summary(model_lmm)
```

@@

# Reshape the dataset to long format
asia_africa_long <- asia_africa %>%
  pivot_longer(cols = c( pro_basicwater, pro_safewater, pro_basicsanitation, 
                        pro_safesanitation,hdi),
               names_to = "Predictor", values_to = "Value")

@@

It can be seen that some variables show u-shaped (quadratic), however some variables looks more complex. In this case, we are going to try to fit two models, first model will fit Mixed effects model with Polynomial Terms, and addictive mixed model (GAMM).

I will perform two steps here: 
1. Data transformation, then looking at possible model, 
2. Directly performing Mixed effects model with polynomial terms, and Performing GAMM

@@

asia_africa <- asia_africa %>%
  mutate(
    log_tbi = log(tbi + 1),
    sqrt_basicwater = sqrt(pmax(pro_basicwater, 0)), 
    sqrt_safewater = sqrt(pmax(pro_safewater, 0)),
    poly_basicsanitation = as.numeric(poly(pro_basicsanitation, 2, raw = TRUE)[, 1]),
    poly_safesanitation = as.numeric(poly(pro_safesanitation, 2, raw = TRUE)[, 1])
  )

@@

```{r}
asia_africa_transformed <- asia_africa %>%
  dplyr::select(log_tbi, poly_basicsanitation, poly_safesanitation,
                sqrt_basicwater, sqrt_safewater, hdi) %>%
  pivot_longer(cols = -log_tbi, names_to = "Predictor", values_to = "Value")

@@

*Generalized Linear mixed model (GLMM)*
After transformation, the data show linear relationship, therefore we perform Linear Mixed Models (LMM) 
```{r}

#Generalized Linear mixed model (GLMM)
#without HDI
library(lme4)
model1<- glmer(log(tbi) ~ sqrt(pro_basicwater) + sqrt(pro_safewater) + 
                          poly(pro_basicsanitation, 2) + poly(pro_safesanitation, 2) +
                          (1 | country), data = asia_africa)
summary(model1)

Interpretation:
✔ Random intercept for country is moderate, meaning tbi varies across countries.
✔ Residual variance is small, indicating that the model explains much of the variability

✔ Intercept (5.24113) is highly significant, representing the baseline log-transformed tbi when all predictors are at their reference values.
✔ poly(pro_basicsanitation, 2)1 (first-order term) is significant, indicating that basic sanitation has a linear effect on tbi.
❌ poly(pro_basicsanitation, 2)2 (second-order term) is not significant, suggesting that the quadratic effect is weak or unnecessary.
❌ pro_basicwater and pro_safewater are not significant, meaning they may not be strong predictors of tbi.

```{r}
#with HDI
library(lme4)
model2<- glmer(log(tbi) ~ sqrt(pro_basicwater) + sqrt(pro_safewater) + 
                          poly(pro_basicsanitation, 2) + poly(pro_safesanitation, 2) +
                          (1 | country) + hdi, data = asia_africa)
summary(model2)

```
Interpretation:

✔ Random intercept for country is moderate, meaning tbi varies across countries.
✔ Residual variance is very small, indicating that the model explains most of the variability, but this may suggest overfitting.

✔ HDI (hdi) is highly significant (p < 0.001), meaning higher HDI is associated with lower tbi.
✔ poly(pro_basicsanitation, 2)1 (first-degree term) is significant, indicating that basic sanitation has a non-linear effect on tbi.
❌ poly(pro_basicsanitation, 2)2 (second-degree term) is not significant, suggesting that the quadratic effect is weak or unnecessary.
❌ pro_basicwater and pro_safewater are not significant, meaning they may not be useful in predicting tbi.

*check wulticolinearity*
```{r}
vif(model_lmm)
```
The Vif test shows that Pro_safesanitation is >5. We need to remove.

*fit model without non-significant predictor*

```{r}
#refit model without non-significant predictor using lmm 
model3 <- glmer(log(tbi) ~ hdi + poly(pro_basicsanitation, 2) + (1 | country), 
                     data = asia_africa)
summary(model3)

```
Interpretation:

The country-level variance (0.42162) is relatively high, indicating that TB incidence varies significantly between countries.
The residual variance (0.01748) is much smaller, meaning a significant portion of TB variation is explained by country-level differences.

Key Findings
✅ HDI is highly significant (Estimate = -3.67, p < 2e-16)

Higher HDI is strongly associated with lower TB incidence.
A 1-unit increase in HDI decreases log(TBI) by ~3.67.
✅ Second-degree polynomial term for sanitation (pro_basicsanitation²) is highly significant (p = 4.15e-06)

Suggests a nonlinear relationship between basic sanitation and TB incidence.
Diminishing or increasing effects at different levels of sanitation.
❌ First-degree polynomial term for sanitation (pro_basicsanitation¹) is NOT significant (p = 0.286)

This indicates that a linear effect of sanitation is weak.
The nonlinear (quadratic) effect is more important.

*Performing Generalized Additive Models (GAM)*

```{r}
model4 <- gamm(log(tbi) ~ s(hdi, bs="ps") + s(pro_basicwater, bs="ps") +
                         s(pro_safewater, bs="ps") + s(pro_basicsanitation, bs="ps") +
                         s(country, bs="re"),
              data = asia_africa_clean, method = "REML")
# bam() from mgcv, which is more efficient for mixed models
summary(model4$gam)
plot(model4$gam)

```{r}
AIC(model1)
AIC(model2)
AIC(model3)
AIC(model4)

BIC(model1)
BIC(model2)
BIC(model3)
BIC(model4)
```
```{r}
# Create a data frame for model comparison
model_comparison <- data.frame(
  Model = c("Model 1: Main Predictors", "Model 2: Including Covariates", "Model 3: Significant variables only", "Model 4: GAMM"),
  AIC = c(AIC(model1), AIC(model2), AIC(model3), AIC(model4)),
  BIC = c(BIC(model1), BIC(model2), BIC(model3), BIC(model4))
)

# Print the table
print(model_comparison)

library(here)

# Define the file location using here()
criteria_location <- here("code", "analysis-code", "tables", "criteria_output.rds")

# Save the model
saveRDS(model_comparison, file = criteria_location)

# Save as a nicely formatted text file
sink("model_comparison.txt")
print(model_comparison)
sink()

Model4 (GAM) Performs better.

Lets visualize model4

# Define the file path using here()
plot_location <- here("code", "analysis-code", "figures", "gam_diagnostics.jpg")

# Open JPEG device
jpeg(plot_location, width = 2000, height = 1500, res = 300)

# Arrange the plots in a grid (2 rows, 3 columns)
par(mfrow = c(2, 3))  
plot(model4$gam, pages = 1)  # Generate all GAM diagnostic plots

# Close the graphics device to save the file
dev.off()

# Confirm file saved
print(paste("Plot saved at:", plot_location))

```{r}
# Create summary table
gam_summary <- summary(model4$gam)

# Extract relevant statistics
gam_table <- data.frame(
  Term = rownames(gam_summary$s.table),
  EDF = gam_summary$s.table[, 1],
  F_Value = gam_summary$s.table[, 3],
  P_Value = gam_summary$s.table[, 4]
)

# Define the file location using here()
summary_location <- here("code", "analysis-code", "tables", "summary_Gamm.rds")

# Save the model
saveRDS(gam_table, file = summary_location)
```

### 2025-03-19 code/eda-code/eda.qmd
Data Analysis (Part 3 of the project)

library(here)
library(knitr)
install.packages("kableExtra")
library(kableExtra)
install.packages("GGally")
library(GGally)
# Install and load the patchwork package
install.packages("patchwork")
library(patchwork)
```

@@

#subset for south east and Asia Africa only 
asia_africa <- mydata %>% 
  filter(region == c("South-East Asia", "Africa"))
unique(asia_africa$country)

@@

# Select only numeric columns (excluding country, year, and region)
asia_africa_long <- asia_africa %>%
  select(-region) %>%  # Exclude categorical variable
  pivot_longer(cols = -c(country, year),  # Convert all numeric variables except country & year
               names_to = "Variable", 
               values_to = "Value")

# Create line plots for all variables
plot2 <- ggplot(asia_africa_long, aes(x = year, y = Value, group = country, color = country)) +
  geom_line(size = 1, alpha = 0.8) +  # Alpha for transparency
  facet_wrap(~ Variable, scales = "free_y") +  # Separate plots for each variable
  labs(title = "Trends of All Variables Over Time",
       x = "Year",
       y = "Value",
       color = "Country") +
  theme_minimal() +
  theme(legend.position = "none",  # Hide legend for clarity
        axis.text.x = element_text(angle = 45, hjust = 1))

@@

plot(plot2)
```

```{r}
plot3 <- ggplot(asia_africa, aes(x = year, y = tbi, group = country, color = country)) +
  geom_line(alpha = 0.7) +  # Transparency to see overlapping lines
  labs(title = "Spaghetti Plot of TBC Incidence",
       x = "Year",
       y = "TBC Incidence") +
  theme_minimal() +
  theme(legend.position = "none")

plot_spageti_all <- here( "code","eda-code", "figures", "spageti-tbi.png") # to set up location for the pictures created 
ggsave(filename = plot_spageti_all, plot=plot3, width = 16, height = 12, units = "in", dpi = 300) # save the pictures created

@@

```{r}
# Find a scaling factor to adjust TBI to match percentage values
max_indicator <- max(asia_africa[, c("pro_safesanitation", "pro_basicsanitation", "pro_safewater", "pro_basicwater", "pro_cleanfuel")], na.rm = TRUE)
max_tbi <- max(asia_africa$tbi, na.rm = TRUE)
scale_factor <- max_indicator / max_tbi  # Scale TBI to fit

# Reshape the data from wide to long format
asia_africa_long <- asia_africa %>%
  pivot_longer(cols = c(pro_safesanitation, pro_basicsanitation, pro_safewater, pro_basicwater, pro_cleanfuel),
               names_to = "Indicator", 
               values_to = "Value")

# Create the plot
plot4 <- ggplot() +
  # First, plot water, sanitation, and clean fuel indicators
  geom_line(data = asia_africa_long, 
            aes(x = year, y = Value, color = Indicator, group = Indicator), 
            size = 1) +
  # Add TBI as a separate line with a different scale
  geom_line(data = asia_africa, 
            aes(x = year, y = tbi * scale_factor, color = "TBI"), 
            size = 1, linetype = "dashed") + 
  facet_wrap(~ country) +  # Separate plots for each country
  theme_minimal() +
  labs(title = "Percentage of Population with Access to Water, Sanitation, Clean Fuel & TBI Trends",
       x = "Year",
       y = "Percentage (%)",
       color = "Indicator") +  # Label for legend
  scale_y_continuous(sec.axis = sec_axis(~ . / scale_factor, name = "TBI")) +  # Secondary Y-axis for TBI
  scale_color_manual(values = c("blue", "green", "purple", "orange", "brown", "red")) +  # Custom colors
  theme(legend.position = "bottom",  
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.border = element_rect(color = "black", fill = NA, size = 1))  # Rotate x-axis labels for readability

@@

```{r}
# Load required libraries
library(dplyr)
library(tidyr)
library(kableExtra)
library(here)

# Define row colors for each category
row_colors <- c(
  "TBI" = "#FFD700", # Gold
  "GDP" = "#ADD8E6", # Light Blue
  "HDI" = "#90EE90", # Light Green
  "Clean Fuel" = "#FFB6C1", # Light Pink
  "Safe Water" = "#87CEEB", # Sky Blue
  "Basic Sanitation" = "#FFA07A", # Light Salmon
  "Safe Sanitation" = "#F08080", # Light Coral
  "PM2.5" = "#D3D3D3", # Light Gray
  "Greenhouse" = "#B0C4DE" # Light Steel Blue
)

# Create summary per year
summary_table <- asia_africa %>%
  group_by(year) %>%  
  summarise(
    Min_TBI = min(tbi, na.rm = TRUE),
    Mean_TBI = mean(tbi, na.rm = TRUE),
    Max_TBI = max(tbi, na.rm = TRUE),
    SD_TBI = sd(tbi, na.rm = TRUE),

    Min_GDP = min(gdp, na.rm = TRUE),
    Mean_GDP = mean(gdp, na.rm = TRUE),
    Max_GDP = max(gdp, na.rm = TRUE),
    SD_GDP = sd(gdp, na.rm = TRUE),

    Min_HDI = min(hdi, na.rm = TRUE),
    Mean_HDI = mean(hdi, na.rm = TRUE),
    Max_HDI = max(hdi, na.rm = TRUE),
    SD_HDI = sd(hdi, na.rm = TRUE),

    Min_CleanFuel = min(pro_cleanfuel, na.rm = TRUE),
    Mean_CleanFuel = mean(pro_cleanfuel, na.rm = TRUE),
    Max_CleanFuel = max(pro_cleanfuel, na.rm = TRUE),
    SD_CleanFuel = sd(pro_cleanfuel, na.rm = TRUE),

    Min_SafeWater = min(pro_safewater, na.rm = TRUE),
    Mean_SafeWater = mean(pro_safewater, na.rm = TRUE),
    Max_SafeWater = max(pro_safewater, na.rm = TRUE),
    SD_SafeWater = sd(pro_safewater, na.rm = TRUE),

    Min_BasicSanitation = min(pro_basicsanitation, na.rm = TRUE),
    Mean_BasicSanitation = mean(pro_basicsanitation, na.rm = TRUE),
    Max_BasicSanitation = max(pro_basicsanitation, na.rm = TRUE),
    SD_BasicSanitation = sd(pro_basicsanitation, na.rm = TRUE),

    Min_SafeSanitation = min(pro_safesanitation, na.rm = TRUE),
    Mean_SafeSanitation = mean(pro_safesanitation, na.rm = TRUE),
    Max_SafeSanitation = max(pro_safesanitation, na.rm = TRUE),
    SD_SafeSanitation = sd(pro_safesanitation, na.rm = TRUE),

    Min_PM2.5 = min(mean_exposure_pm2.5, na.rm = TRUE),
    Mean_PM2.5 = mean(mean_exposure_pm2.5, na.rm = TRUE),
    Max_PM2.5 = max(mean_exposure_pm2.5, na.rm = TRUE),
    SD_PM2.5 = sd(mean_exposure_pm2.5, na.rm = TRUE),

    Min_Greenhouse = min(total_greenhouses, na.rm = TRUE),
    Mean_Greenhouse = mean(total_greenhouses, na.rm = TRUE),
    Max_Greenhouse = max(total_greenhouses, na.rm = TRUE),
    SD_Greenhouse = sd(total_greenhouses, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = -year, names_to = "Statistic", values_to = "Value") %>%
  pivot_wider(names_from = year, values_from = Value) %>% # Convert to wide format
  rowwise() %>% 
  mutate(Accumulated = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(across(where(is.numeric), ~ sprintf("%.2f", .x))) %>% # Format numbers to 2 decimal places
  relocate(Accumulated, .after = Statistic) %>% # Move Accumulated after Statistic
  mutate(Row_Index = row_number(), # Create row indices for coloring
         Category = case_when(
           grepl("TBI", Statistic) ~ "TBI",
           grepl("GDP", Statistic) ~ "GDP",
           grepl("HDI", Statistic) ~ "HDI",
           grepl("CleanFuel", Statistic) ~ "Clean Fuel",
           grepl("SafeWater", Statistic) ~ "Safe Water",
           grepl("BasicSanitation", Statistic) ~ "Basic Sanitation",
           grepl("SafeSanitation", Statistic) ~ "Safe Sanitation",
           grepl("PM2.5", Statistic) ~ "PM2.5",
           grepl("Greenhouse", Statistic) ~ "Greenhouse"
         ))

# Generate table with colors
summary_table_display <- summary_table %>%
  select(-Category, -Row_Index) %>% # Remove Category before displaying
  kable(caption = "Summary Statistics by Year with Accumulated Column") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

# Apply row coloring dynamically
for (cat in names(row_colors)) {
  rows_to_color <- which(summary_table$Category == cat) # Get row indices
  summary_table_display <- summary_table_display %>%
    row_spec(rows_to_color, background = row_colors[cat])
}

saveRDS(summary_table_display, file = here("code", "eda-code", "tables", "summary_table.html"))
print(summary_table_display)

@@

```{r}
#| label: tbl-resulttable1
#| tbl-cap: "Summary Statistics Table"
#| echo: FALSE

htmltools::includeHTML(here("code", "eda-code", "tables", "summary_table.html"))

@@

```{r}
gg1 <- ggplot(asia_africa, aes(x = gdp, y = tbi)) +
  geom_point() +  # Create scatter plot
  ggtitle("Scatter Plot: tbi vs GDP") +  # Add title
  xlab("GDP") +  # Label the x-axis
  ylab("TBI") +  # Label the y-axis
  theme_minimal()

gg2 <- 
```

@@

# Create a list of scatter plots for each independent variable against tbi
plots <- lapply(independent_vars, function(var) {
  ggplot(asia_africa, aes_string(x = var, y = "tbi")) +
    geom_point() + 
    ggtitle(paste("tbi vs", var)) + 
    xlab(var) + 
    ylab("TBI") + 
    theme_minimal()
})

# Combine all plots into a single frame
combined_plot <- wrap_plots(plots, ncol = 3)  # Adjust ncol to the number of columns you want
combined_plot

combineplot_location <- here( "code","eda-code", "figures", "scatterplot.png") # to set up location for the pictures created 
ggsave(filename = combineplot_location, plot=combined_plot, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
plot(combined_plot)

>>>>

library(here)
library(knitr)
install.packages("kableExtra")
library(kableExtra)
install.packages("GGally")
library(GGally)

# Install and load the patchwork package
install.packages("patchwork")
library(patchwork)

install.packages("gtsummary")
library(gtsummary)
```

@@

#subset for south east and Asia Africa only 
asia_africa <- mydata %>% 
  filter(region == c("South-East Asia", "Africa"))

asia_africa <- asia_africa%>%
  filter(country != "Lesotho") #Drop Lesotho from the observation

unique(asia_africa$country)

@@

# Reshape data to long format for faceting
asia_africa_long <- asia_africa %>%
  pivot_longer(cols = c(tbi, pro_basicwater, pro_safewater, pro_basicsanitation, 
                        pro_safesanitation, gdp, hdi),
               names_to = "variable", values_to = "value")

# Create spaghetti plot with facets
plot2 <- ggplot(asia_africa_long, aes(x = year, y = value, color = country)) + 
  geom_line(aes(group = country), alpha = 0.7) +  # Different color for each country
  stat_smooth(method = "loess", col = "black", lwd = 1.5, se = FALSE) +  # Mean line in black
  scale_x_continuous(name = "Year") + 
  scale_y_continuous(name = "Value") + 
  facet_wrap(~ variable, scales = "free_y") +  # Create facets for each variable
  theme_bw() + 
  theme(axis.text = element_text(size = 12, colour = "black"), 
        axis.title = element_text(size = 14, face = "bold"),
        strip.text = element_text(size = 14, face = "bold"),  # Style facet titles
        legend.position = "none")  # Remove legend if not needed

@@

plot(plot2)

plot3 <-  ggplot(asia_africa, aes(x = year, y = tbi, color = country)) + 
  geom_line(aes(group = country), alpha = 0.7) +  # Different color for each country
  stat_smooth(method = "loess", col = "black", lwd = 1.5, se = FALSE) +  # Mean line in black
  scale_x_continuous(name = "Year") + 
  scale_y_continuous(name = "Tuberculosis Incidence") +
  theme_bw() + 
  theme(axis.text = element_text(size = 14, colour = "black"), 
        axis.title = element_text(size = 14, face = "bold"),
        strip.text.x = element_text(size = 14),
        legend.position = "none")  # Remove legend if not needed
plot_spageti_all <- here( "code","eda-code", "figures", "spageti-tbi.png") # to set up location for the pictures created 
ggsave(filename = plot_spageti_all, plot=plot3, width = 16, height = 12, units = "in", dpi = 300) # save the pictures created

@@

# Find a scaling factor to adjust TBI to match percentage values
max_indicator <- max(asia_africa[, c("pro_safesanitation", "pro_basicsanitation", "pro_safewater", "pro_basicwater")], na.rm = TRUE)
max_tbi <- max(asia_africa$tbi, na.rm = TRUE)
scale_factor <- max_indicator / max_tbi  # Scale TBI to fit

# Reshape the data from wide to long format
asia_africa_long <- asia_africa %>%
  pivot_longer(cols = c(pro_safesanitation, pro_basicsanitation, pro_safewater, pro_basicwater, tbi),
               names_to = "Indicator", 
               values_to = "Value")

# Create the plot
plot4 <- ggplot() +
  # First, plot water, sanitation, and clean fuel indicators
  geom_line(data = asia_africa_long, 
            aes(x = year, y = Value, color = Indicator, group = Indicator), 
            size = 1) +
  # Add TBI as a separate line with a different scale
  geom_line(data = asia_africa, 
            aes(x = year, y = tbi * scale_factor, color = "tbi"), 
            size = 1, linetype = "dashed") + 
  facet_wrap(~ country) +  # Separate plots for each country
  theme_minimal() +
  labs(title = "Percentage of Population with Access to Water, Sanitation & TBI Trends",
       x = "Year",
       y = "Percentage (%)",
       color = "Indicator") +  # Label for legend
  scale_y_continuous(sec.axis = sec_axis(~ . / scale_factor, name = "TBI")) +  # Secondary Y-axis for TBI
  scale_color_manual(values = c("blue", "green", "purple", "orange", "red")) +  # Custom colors
  theme(legend.position = "bottom",  
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.border = element_rect(color = "black", fill = NA, size = 1))  # Rotate x-axis labels for readability

@@

summary_table <- tbl_summary(asia_africa, include = -c(country, region))

saveRDS(summary_table, file = here("code", "eda-code", "tables", "summary_table.rds"))
print(summary_table)

@@

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable= read_rds(here("code", "eda-code", "tables", "summary_table.rds"))
knitr::kable(resulttable, format = "pipe")

@@

@@

library(ggplot2)
library(dplyr)
library(tidyr)

# Convert dataset to long format for faceting
asia_africa_long <- asia_africa %>%
  pivot_longer(cols = c( pro_basicwater, pro_safewater, 
                        pro_basicsanitation, pro_safesanitation, hdi),
               names_to = "variable", values_to = "value")

# Create scatter plot faceted by independent variable
scatter_faceted <- ggplot(asia_africa_long, aes(x = value, y = tbi, color = country)) + 
  geom_point(alpha = 0.6) +  # Scatter points, colored by country
  stat_smooth(method = "loess", col = "black", lwd = 1.2, se = FALSE) +  # LOESS (non-linear) in black
  facet_wrap(~ variable, scales = "free_x") +  # Facet by variable
  labs(x = "Independent Variable Value", y = "Tuberculosis Incidence (TBI)") +
  theme_bw() + 
  theme(axis.text = element_text(size = 12, colour = "black"), 
        axis.title = element_text(size = 14, face = "bold"),
        strip.text = element_text(size = 14, face = "bold"), 
        legend.position = "none")  # Hides the legend

# Show plot
print(scatter_faceted)

library(ggplot2)
library(patchwork)  # For arranging multiple plots

# Define independent variables
independent_vars <- c("hdi", "pro_basicsanitation", "pro_safesanitation", 
                      "pro_basicwater", "pro_safewater")

# Create individual scatter plots
plots <- lapply(independent_vars, function(var) {
  ggplot(asia_africa, aes_string(x = var, y = "tbi")) +  
    geom_point(color = "black") +  # Black points only, no line
    ggtitle(paste("TBI vs", var)) + 
    xlab(var) + 
    ylab("TBI") + 
    theme_minimal()
})

combineplot_location <- here( "code","eda-code", "figures", "scatterplot.png") # to set up location for the pictures created 
ggsave(filename = combineplot_location, plot=final_plot, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
plot(final_plot)

```{r}
# Remove rows where tbi or any independent variable has missing values
asia_africa_clean <- asia_africa %>%
  filter(!is.na(tbi)) %>% # Remove missing tbi values
  drop_na(all_of(independent_vars))  # Remove missing values in independent variables

# Create scatter plots again with cleaned data
plots <- lapply(independent_vars, function(var) {
  ggplot(asia_africa_clean, aes_string(x = var, y = "tbi", color = "country")) +
    geom_point(alpha = 0.7) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "solid", size = 1) +
    ggtitle(paste("TBI vs", var)) + 
    xlab(var) + 
    ylab("TBI") + 
    theme_minimal() +
    theme(legend.position = "none")  
})

# Combine all plots
combined_plot <- wrap_plots(plots, ncol = 3)

# Save the plot
#combineplot_location <- here("code", "eda-code", "figures", "scatterplot.png") 
#ggsave(filename = combineplot_location, plot = combined_plot, width = 16, height = 12, units = "in", dpi = 300)

# Display the plot
plot(combined_plot)
